---
title: "NBA Analysis - Project 1"
author: "Tarun"
date: "2026-05-14"
output: html_document
---


```{r}

knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 6
)



# Load libraries
library(tidyverse)
library(ggplot2)
library(knitr)

```


---

## 1. Project Overview

This analysis explores the **scoring efficiency** of 10 elite NBA players between the
2005 and 2014 seasons. Using data on points scored, minutes played, field goals made,
and field goal attempts, we answer three key questions:

1. Which player scored the most points **per minute** played?
2. Which player had the highest **field goal percentage** (shooting accuracy)?
3. How did each player's **efficiency change over time**?

---

## 2. Loading the Data

We begin by sourcing the basketball dataset, which contains six matrices covering
salary, games played, minutes played, field goals, field goal attempts, and points.

```{r}
file.exists("data/basketball_data.R")

```


```{r load_data}

# Source the data file — make sure basketball_data.R is in your /data folder
source("data/basketball_data.R")


# Quick check — confirm all matrices are loaded
cat("Players:", Players, "\n")
cat("Seasons:", Seasons, "\n")
cat("Dimensions of Points matrix:", dim(Points), "\n")
```

---




## 3. Data Wrangling

We convert the raw matrices into a tidy data frame and calculate two key efficiency
metrics:

- **Points Per Minute (PPM)** — how many points a player scores per minute on court
- **Field Goal Percentage (FG%)** — proportion of shot attempts that are successful

```{r wrangle_data}
# Convert matrices to tidy data frames
points_df <- as.data.frame(Points) |>
  rownames_to_column("player") |>
  pivot_longer(-player, names_to = "season", values_to = "points")

minutes_df <- as.data.frame(MinutesPlayed) |>
  rownames_to_column("player") |>
  pivot_longer(-player, names_to = "season", values_to = "minutes")

fg_df <- as.data.frame(FieldGoals) |>
  rownames_to_column("player") |>
  pivot_longer(-player, names_to = "season", values_to = "fg_made")

fga_df <- as.data.frame(FieldGoalAttempts) |>
  rownames_to_column("player") |>
  pivot_longer(-player, names_to = "season", values_to = "fg_attempts")

games_df <- as.data.frame(Games) |>
  rownames_to_column("player") |>
  pivot_longer(-player, names_to = "season", values_to = "games")

# Join all into one master data frame
nba_df <- points_df |>
  left_join(minutes_df,  by = c("player", "season")) |>
  left_join(fg_df,       by = c("player", "season")) |>
  left_join(fga_df,      by = c("player", "season")) |>
  left_join(games_df,    by = c("player", "season")) |>
  mutate(
    season        = as.integer(season),
    # Points per minute — exclude seasons with 0 minutes (injury/no play)
    ppm           = ifelse(minutes > 0, round(points / minutes, 4), NA),
    # Field goal percentage
    fg_pct        = ifelse(fg_attempts > 0, round(fg_made / fg_attempts * 100, 2), NA),
    # Points per game
    ppg           = ifelse(games > 0, round(points / games, 2), NA),
    # Clean player names for display
    player_clean  = str_replace_all(player, "([A-Z])", " \\1") |> str_trim()
  )

# Preview the cleaned data
head(nba_df, 10) |>
  select(player_clean, season, points, minutes, ppm, fg_pct, ppg) |>
  kable(caption = "Sample of cleaned NBA data", col.names = c("Player","Season","Points","Minutes","Pts/Min","FG%","Pts/Game"))
```



---

## 4. Overall Career Efficiency

### 4.1 Average Points Per Minute (Career)

Which player made the most of their time on court across all 10 seasons?

```{r ppm_career_bar}
# Calculate career average PPM (excluding injury seasons with 0 minutes)
career_ppm <- nba_df |>
  filter(!is.na(ppm)) |>
  group_by(player_clean) |>
  summarise(avg_ppm = mean(ppm), .groups = "drop") |>
  arrange(desc(avg_ppm))

# Bar chart
ggplot(career_ppm, aes(x = reorder(player_clean, avg_ppm), y = avg_ppm, fill = avg_ppm)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = round(avg_ppm, 3)), hjust = -0.15, size = 3.5) +
  coord_flip() +
  scale_fill_gradient(low = "#f4a460", high = "#c0392b", guide = "none") +
  scale_y_continuous(limits = c(0, 1.1)) +
  labs(
    title    = "Career Average Points Per Minute (2005–2014)",
    subtitle = "Higher = more efficient scorer per minute on court",
    x        = NULL,
    y        = "Points Per Minute"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(colour = "grey50"),
    panel.grid.major.y = element_blank()
  )
```

