---
title: "NBA Analysis - Project 1"
author: "Tarun"
date: "2026-05-14"
output: html_document
---


```{r}

knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 6
)



# Load libraries
library(tidyverse)
library(ggplot2)
library(knitr)

```


---

## 1. Project Overview

This analysis explores the **scoring efficiency** of 10 elite NBA players between the
2005 and 2014 seasons. Using data on points scored, minutes played, field goals made,
and field goal attempts, we answer three key questions:

1. Which player scored the most points **per minute** played?
2. Which player had the highest **field goal percentage** (shooting accuracy)?
3. How did each player's **efficiency change over time**?

---

## 2. Loading the Data

We begin by sourcing the basketball dataset, which contains six matrices covering
salary, games played, minutes played, field goals, field goal attempts, and points.

```{r}
file.exists("data/basketball_data.R")

```


```{r load_data}

# Source the data file — make sure basketball_data.R is in your /data folder
source("data/basketball_data.R")


# Quick check — confirm all matrices are loaded
cat("Players:", Players, "\n")
cat("Seasons:", Seasons, "\n")
cat("Dimensions of Points matrix:", dim(Points), "\n")
```

---
